---
title: "Traceable ADaM derivation: every exclusion documented"
subtitle: "Row-level data provenance for a Phase III atopic dermatitis trial"
author: "Ndoh Penn"
date: "2026-06-26"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Traceable ADaM derivation: every exclusion documented}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(2026)
knitr::opts_chunk$set(comment = "#>", fig.width = 7, fig.height = 4.5)
```

## Overview

Every ADaM derivation involves exclusions — screen failures, protocol
deviations, missing baselines, assessment windows. In most trials these are
documented in aggregate: "47 subjects were excluded from the per-protocol
population." But regulators increasingly ask for more: which rows? Why? And
can you trace any subject through the complete derivation pipeline?

`lineager` answers these questions at the row level. Every row in every
dataset carries a lineage ID (`lineage_id`) that survives filters, joins, and
derivations. Every `lg_filter()` call requires a documented reason, and the
session automatically accumulates an exclusion registry as the pipeline
runs — no manual disposition tracking needed.

This article builds an ADEFF dataset for a Phase III atopic dermatitis
trial — starting from raw ADSL and ADLB, deriving baseline and change from
baseline, tracking every exclusion, and producing a CDISC Reviewer's
Guide-aligned provenance report.

---

## Setup

```{r packages}
library(lineager)
library(dplyr)
```

---

## Simulate source datasets

We simulate CDISC ADaM-structured ADSL (subject-level) and ADLB (lab/
assessment) datasets for a Phase III trial of dupilumab in moderate-to-severe
atopic dermatitis. 320 randomised subjects, primary endpoint EASI score.

```{r simulate}
n_total <- 320L

# ── ADSL ──────────────────────────────────────────────────────────────────
adsl_raw <- data.frame(
  STUDYID  = "DERM-DUP-301",
  USUBJID  = sprintf("DUP301-%04d", seq_len(n_total)),
  TRT01P   = rep(c("Dupilumab 300mg Q2W", "Placebo Q2W"), each = n_total / 2),
  AGE      = round(rnorm(n_total, mean = 36, sd = 14)),
  SEX      = sample(c("M", "F"), n_total, replace = TRUE),
  ITTFL    = ifelse(runif(n_total) < 0.97, "Y", "N"),
  SAFFL    = ifelse(runif(n_total) < 0.98, "Y", "N"),
  PPROTFL  = ifelse(runif(n_total) < 0.91, "Y", "N"),
  EASISCAT = sample(c("MODERATE", "SEVERE"), n_total, replace = TRUE, prob = c(0.55, 0.45)),
  stringsAsFactors = FALSE
)

# ── ADLB: EASI scores at baseline and Week 16 ──────────────────────────────
adlb_raw <- expand.grid(
  USUBJID = adsl_raw$USUBJID,
  AVISITN = c(0L, 16L),
  stringsAsFactors = FALSE
) |>
  merge(adsl_raw[, c("USUBJID", "TRT01P", "EASISCAT")], by = "USUBJID") |>
  arrange(USUBJID, AVISITN)

set.seed(2026)
adlb_raw$AVAL <- mapply(function(trt, vis, scat) {
  base_mean <- if (scat == "SEVERE") 45 else 30
  if (vis == 0L) return(round(pmax(0, rnorm(1, base_mean, 8)), 1))
  reduction <- if (trt == "Dupilumab 300mg Q2W") 0.55 else 0.12
  round(pmax(0, base_mean * (1 - reduction) + rnorm(1, 0, 5)), 1)
}, adlb_raw$TRT01P, adlb_raw$AVISITN, adlb_raw$EASISCAT)

# Missingness at Week 16 (~6%)
post_bl <- adlb_raw$AVISITN > 0
miss_idx <- which(post_bl)[sample(sum(post_bl), round(sum(post_bl) * 0.06))]
adlb_raw$AVAL[miss_idx] <- NA_real_

adlb_raw$PARAMCD <- "EASI"

cat(sprintf(
  "ADSL: %d subjects | ADLB: %d records\n", nrow(adsl_raw), nrow(adlb_raw)
))
```

---

## Start the lineager session

`lg_start()` initialises the session store. All `lg_*` functions write to
this store automatically.

```{r session}
lg_start(study_id = "DERM-DUP-301", analysis_id = "ADEFF-primary-efficacy")
```

---

## Tag source datasets

`lg_tag()` assigns a unique lineage ID (`lineage_id`) to every row. The
`USUBJID` is embedded automatically when present, making IDs human-readable.

```{r tag}
adsl <- lg_tag(adsl_raw, dataset_id = "ADSL", domain = "DM", label = "Subject-Level")
adlb <- lg_tag(adlb_raw, dataset_id = "ADLB", domain = "LB", label = "Laboratory/Efficacy")

head(adsl[, c("lineage_id", "USUBJID", "TRT01P", "ITTFL")], 4L)
```

---

## Build the ADEFF derivation pipeline

### Step 1: Document and apply the safety population

`lg_population()` registers the population flag definition — inclusion
criteria, exclusion criteria, and a plain-English description — directly
linking to the flag already present in the data.

```{r population}
lg_population(
  adsl,
  flag_var      = "SAFFL",
  label         = "Safety Analysis Flag",
  definition    = "All randomised subjects who received at least one dose of study treatment",
  incl_criteria = c("Randomised", "Received >= 1 dose of study treatment"),
  excl_criteria = "No study drug administered"
)

adsl_saf <- lg_filter(
  adsl, SAFFL == "Y",
  reason     = "Restrict to safety analysis set (SAFFL = Y) per SAP Section 4.1",
  population = "SAFFL"
)
```

### Step 2: ITT population

```{r itt}
adsl_itt <- lg_filter(
  adsl_saf, ITTFL == "Y",
  reason     = "Restrict to ITT population (ITTFL = Y) per SAP Section 4.2",
  population = "ITTFL"
)

cat(sprintf("After ITT filter: %d subjects\n", nrow(adsl_itt)))
```

### Step 3: Join EASI assessments to ADSL covariates

```{r join}
adeff <- lg_join(
  adlb, adsl_itt[, c("USUBJID", "TRT01P", "EASISCAT")],
  by          = "USUBJID",
  type        = "inner",
  description = "Join ADLB EASI assessments to ITT-restricted ADSL covariates"
)

cat(sprintf("After join: %d records\n", nrow(adeff)))
```

### Step 4: Derive baseline and change from baseline

```{r derive-baseline}
adeff <- adeff |>
  group_by(USUBJID) |>
  mutate(BASE = AVAL[AVISITN == 0L][1L]) |>
  ungroup()

# Re-tag after the dplyr grouping pipeline to keep lg_df class intact
# Drop any existing lineage_id before re-tagging to avoid duplicate columns
adeff_df <- as.data.frame(adeff)
adeff_df <- adeff_df[, !names(adeff_df) %in% "lineage_id"]
adeff <- lg_tag(adeff_df, dataset_id = "ADEFF", domain = "ADLB",
                label = "Efficacy Analysis Dataset")

n_miss_base <- sum(is.na(adeff$BASE[adeff$AVISITN == 0L]))

adeff <- lg_filter(
  adeff, !is.na(BASE),
  reason = "Exclude records with missing baseline EASI score per SAP Section 5.1",
  population = "ITTFL"
)

adeff <- lg_derive(
  adeff,
  CHG  = AVAL - BASE,
  PCHG = (AVAL - BASE) / BASE * 100,
  description = "Derive change from baseline (CHG) and percent change (PCHG) per CDISC ADaM IG Section 3.2.7"
)
```

### Step 5: Restrict to Week 16, non-missing

```{r week16}
adeff_w16 <- lg_filter(
  adeff, AVISITN == 16L, !is.na(AVAL),
  reason     = "Restrict to Week 16 primary timepoint with non-missing EASI assessment",
  population = "ITTFL"
)

cat(sprintf("Week 16 primary analysis records: %d\n", nrow(adeff_w16)))
```

### Step 6: Derive EASI-75 responder flag

```{r easi75}
adeff_final <- lg_derive(
  adeff_w16,
  EASI75FL = ifelse(PCHG <= -75, "Y", "N"),
  description = "Derive EASI-75 responder flag: Y if percent change <= -75% per SAP Section 5.3"
)
```

---

## Document the SDTM-to-ADaM derivation

`lg_spec()` records the structured derivation specification linking the
ADaM variable back to its SDTM source — the basis for the CDISC Reviewer's
Guide derivation section.

```{r spec}
lg_spec(
  adam_dataset  = "ADEFF",
  adam_var      = "CHG",
  label         = "Change from Baseline EASI Score",
  source_domain = "LB",
  source_var    = "LBSTRESN",
  derivation    = "AVAL - BASE, where BASE is the Visit 0 (Baseline) EASI assessment",
  conditions    = "Subjects with non-missing baseline EASI score"
)
```

---

## Subject disposition

`lg_disposition()` summarises the exclusion registry automatically — no
manual group definitions needed. Every `lg_filter()` call above contributed
its exclusions to this summary.

```{r disposition}
lg_disposition(by = "reason")
```

---

## Exclusion registry

```{r exclusions}
excl <- lg_exclusions()

excl |>
  select(dataset_id, reason, population) |>
  knitr::kable(caption = "ADEFF derivation exclusion registry — DERM-DUP-301")
```

---

## Trace a subject through the pipeline

`lg_trace()` returns the complete provenance history for any subject —
every dataset it appeared in, every operation that touched it, and any
exclusion records.

```{r trace}
example_subj <- adeff_final$USUBJID[1L]
lg_trace(example_subj)
```

```{r trace-excluded}
excluded_subj <- adsl_raw$USUBJID[adsl_raw$ITTFL == "N"][1L]
if (!is.na(excluded_subj)) lg_trace(excluded_subj)
```

---

## Pipeline lineage graph

```{r lineage}
lin <- lg_lineage()
print(lin)
```

---

## Provenance report

```{r report, eval=FALSE}
lg_report(
  output   = "outputs/DERM-DUP-301_ADEFF_provenance_v1.html",
  title    = "ADEFF Provenance Report — DERM-DUP-301",
  study_id = "DERM-DUP-301",
  sponsor  = "Example Pharma Ltd",
  author   = "J. Smith, Biostatistician"
)
```

---

## End the session

```{r end}
lg_end()
```

---

## What the provenance record proves

Every row in the final ADEFF dataset carries a lineage ID tracing it to its
source record in ADSL or ADLB. The exclusion registry — built automatically
from each `lg_filter()` call, not maintained separately — documents which
subjects were excluded at each step, why, and which population the
exclusion relates to. `lg_disposition()` produces the population flow table
directly from these records, so the numbers can never drift out of sync
with the derivation code that produced them.

---

## Session information

```{r session-info}
sessionInfo()
```
