---
title: "Row-level provenance in real-world evidence studies"
subtitle: "New-user cohort derivation with complete exclusion documentation using lineager"
author: "Ndoh Penn"
date: "2026-06-26"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Row-level provenance in real-world evidence studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(2026)
knitr::opts_chunk$set(comment = "#>", fig.width = 7, fig.height = 4.5)
```

## Overview

Real-world evidence studies live or die by their cohort definition. The
new-user design, washout periods, lookback windows, prevalent user
exclusions, and outcome ascertainment rules are all critical analytical
decisions that HEOR reviewers, payers, and regulators scrutinise. Yet in
most RWE studies, the documentation of these decisions is at the aggregate
level: a CONSORT-style flow chart with total numbers at each step.

`lineager` brings row-level traceability to RWE cohort derivation. Every
patient record carries a lineage ID. Every `lg_filter()` call requires a
documented reason, and the session accumulates an exclusion registry
automatically as the pipeline runs. Any patient can be traced through the
complete cohort derivation — from raw claims to the final analysis cohort —
at any time.

This article demonstrates a complete new-user cohort derivation for a
retrospective comparative effectiveness study of two diabetes medications
in administrative claims data.

---

## The study

A retrospective new-user active comparator study comparing GLP-1 receptor
agonists (GLP-1 RA) versus DPP-4 inhibitors in adult patients with type 2
diabetes newly initiating therapy. Primary outcome: composite of major
adverse cardiovascular events (MACE) at 12 months.

---

## Setup

```{r packages}
library(lineager)
library(dplyr)
library(ggplot2)
```

---

## Simulate administrative claims data

```{r simulate}
n_pool <- 25000L  # Patient pool before eligibility screening

set.seed(2026)
claims_raw <- data.frame(
  PATID       = sprintf("PT-%07d", seq_len(n_pool)),
  INDEX_DATE  = as.Date("2022-01-01") +
                  sample(0:729, n_pool, replace = TRUE),
  AGE_INDEX   = round(rnorm(n_pool, mean = 61, sd = 12)),
  SEX         = sample(c("M", "F"), n_pool, replace = TRUE, prob = c(0.48, 0.52)),
  INDEX_DRUG  = sample(c("GLP1RA", "DPP4i"), n_pool, replace = TRUE,
                       prob = c(0.42, 0.58)),
  LOOKBACK_MO = sample(6:36, n_pool, replace = TRUE),
  T2DM_DX     = sample(c(TRUE, FALSE), n_pool, replace = TRUE, prob = c(0.82, 0.18)),
  PRIOR_USE   = sample(c(TRUE, FALSE), n_pool, replace = TRUE, prob = c(0.28, 0.72)),
  INSULIN_USE = sample(c(TRUE, FALSE), n_pool, replace = TRUE, prob = c(0.18, 0.82)),
  CKD_STAGE   = sample(c("NONE", "1-2", "3", "4", "5"),
                       n_pool, replace = TRUE,
                       prob = c(0.55, 0.18, 0.16, 0.08, 0.03)),
  HF_DX       = sample(c(TRUE, FALSE), n_pool, replace = TRUE, prob = c(0.14, 0.86)),
  ESRD        = sample(c(TRUE, FALSE), n_pool, replace = TRUE, prob = c(0.04, 0.96)),
  FOLLOW_MO   = round(runif(n_pool, min = 0.5, max = 24), 1),
  MACE_12M    = sample(c(TRUE, FALSE), n_pool, replace = TRUE, prob = c(0.07, 0.93)),
  stringsAsFactors = FALSE
)

claims_raw$MACE_12M[claims_raw$FOLLOW_MO < 12] <- FALSE
claims_raw$ESRD[claims_raw$CKD_STAGE != "5"] <- FALSE

cat(sprintf("Claims pool: %d patients\n", nrow(claims_raw)))
```

---

## Start the lineager session

```{r session}
lg_start(study_id = "T2DM-GLP1-RWE-001", analysis_id = "new-user-cohort-derivation")
```

---

## Tag the source dataset

`lg_tag()` embeds a column named `USUBJID` into the lineage ID, when one is
present, for human-readable IDs. Administrative claims data conventionally
uses a different identifier (`PATID` here) — so we add a `USUBJID` column
purely for lineage embedding, while `PATID` remains the identifier used
throughout the actual analysis code.

```{r tag}
claims_raw$USUBJID <- claims_raw$PATID

claims <- lg_tag(claims_raw, dataset_id = "CLAIMS_RAW", label = "Raw claims extract")

cat(sprintf("Tagged %d patient records\n", nrow(claims)))
head(claims[, c(".__lid__", "PATID", "INDEX_DRUG", "AGE_INDEX")], 4L)
```

---

## New-user cohort derivation

Each eligibility criterion is a separate `lg_filter()` call with a
documented reason. The exclusion registry and operation log accumulate
automatically — there is no separate disposition tracking to maintain.

### Criterion 1: Age 40–89 at index date

```{r c1}
cohort <- lg_filter(
  claims,
  AGE_INDEX >= 40L & AGE_INDEX <= 89L,
  reason     = "Eligible age range 40-89 years at index date per SAP Section 3.1. Excludes paediatric patients and those with extreme age where T2DM prevalence and drug indication differ substantially.",
  population = "AGE_ELIGIBLE"
)
cat(sprintf("After age criterion: %d patients\n", nrow(cohort)))
```

### Criterion 2: Type 2 diabetes diagnosis

```{r c2}
cohort <- lg_filter(
  cohort, T2DM_DX == TRUE,
  reason     = "Require confirmed T2DM diagnosis (ICD-10 E11.x) in 12 months prior to index date per SAP Section 3.1.2.",
  population = "T2DM_CONFIRMED"
)
cat(sprintf("After T2DM diagnosis criterion: %d patients\n", nrow(cohort)))
```

### Criterion 3: Minimum 12-month lookback

```{r c3}
cohort <- lg_filter(
  cohort, LOOKBACK_MO >= 12L,
  reason     = "Require >= 12 months continuous enrolment prior to index date per SAP Section 3.2.",
  population = "LOOKBACK_ADEQUATE"
)
cat(sprintf("After lookback criterion: %d patients\n", nrow(cohort)))
```

### Criterion 4: New user — no prior use of index drug class

```{r c4}
cohort <- lg_filter(
  cohort, PRIOR_USE == FALSE,
  reason     = "Exclude prevalent users: no dispensing of index drug class in the 12-month lookback period per SAP Section 3.3 (new-user design).",
  population = "NEW_USER"
)
cat(sprintf("After new-user criterion: %d patients\n", nrow(cohort)))
```

### Criterion 5: Exclude insulin users

```{r c5}
cohort <- lg_filter(
  cohort, INSULIN_USE == FALSE,
  reason     = "Exclude patients on insulin at index date per SAP Section 3.4.",
  population = "NO_INSULIN"
)
cat(sprintf("After insulin exclusion: %d patients\n", nrow(cohort)))
```

### Criterion 6: Exclude end-stage renal disease

```{r c6}
cohort <- lg_filter(
  cohort, ESRD == FALSE,
  reason     = "Exclude ESRD (CKD Stage 5 or renal replacement therapy) at or before index date per SAP Section 3.5.",
  population = "NO_ESRD"
)
cat(sprintf("After ESRD exclusion: %d patients\n", nrow(cohort)))
```

### Criterion 7: Minimum 12 months follow-up available

```{r c7}
cohort <- lg_filter(
  cohort, FOLLOW_MO >= 12L,
  reason     = "Require >= 12 months follow-up from index date for the primary 12-month MACE outcome per SAP Section 3.6.",
  population = "FOLLOWUP_ADEQUATE"
)
cat(sprintf("After follow-up criterion: %d patients\n", nrow(cohort)))
```

---

## Document the analysis population

`lg_population()` registers the cumulative cohort definition as a named
population flag, linking the SAP criteria to a single documented object.

```{r population}
cohort$NUC_FL <- "Y"  # all rows remaining at this point are the final cohort

lg_population(
  cohort,
  flag_var      = "NUC_FL",
  label         = "New-User Cohort Flag",
  definition    = "Adults aged 40-89 with confirmed T2DM, adequate lookback, new to index drug class, no insulin use, no ESRD, and adequate follow-up",
  incl_criteria = c("Age 40-89", "T2DM confirmed", ">=12mo lookback",
                    "New user of index drug", "No insulin", "No ESRD",
                    ">=12mo follow-up"),
  excl_criteria = "Any of the above criteria not met"
)
```

---

## Derive analysis variables

```{r derive-vars}
cohort <- lg_derive(
  cohort,
  CKD_MOD_SEV  = CKD_STAGE %in% c("3", "4"),
  HF_FLAG      = HF_DX == TRUE,
  HIGH_RISK_CV = (AGE_INDEX >= 65L) | (HF_DX == TRUE) | (CKD_STAGE %in% c("3","4","5")),
  description  = "Derive cardiovascular risk flags for propensity score model: CKD_MOD_SEV (Stage 3/4), HF_FLAG, composite HIGH_RISK_CV per SAP Section 5.2"
)

cohort <- lg_derive(
  cohort,
  STUDY_YEAR = as.integer(format(INDEX_DATE, "%Y")),
  description = "Derive index year for calendar time confounding adjustment"
)

cohort_final <- lg_derive(
  cohort,
  MACE_FLAG = MACE_12M == TRUE,
  description = "Derive primary outcome indicator from composite MACE endpoint at 12 months per SAP Section 4.1"
)
```

---

## Final cohort composition

```{r cohort-summary}
cohort_char <- cohort_final |>
  group_by(INDEX_DRUG) |>
  summarise(
    N            = n(),
    Age_mean     = round(mean(AGE_INDEX), 1),
    Female_pct   = round(mean(SEX == "F") * 100, 1),
    HighRisk_pct = round(mean(HIGH_RISK_CV) * 100, 1),
    MACE_12m_pct = round(mean(MACE_FLAG) * 100, 1),
    .groups      = "drop"
  )

knitr::kable(
  cohort_char,
  col.names = c("Treatment", "N", "Age (mean)", "Female (%)",
                "High CV risk (%)", "MACE 12m (%)"),
  caption   = "Final cohort characteristics — T2DM-GLP1-RWE-001"
)
```

---

## Exclusion registry and attrition

`lg_exclusions()` returns the row-level exclusion registry. `lg_operations()`
returns the operation-level log with `rows_in`/`rows_out` at each step —
the right source for an attrition waterfall.

```{r exclusion-registry}
excl <- lg_exclusions()

excl |>
  count(population, reason) |>
  knitr::kable(
    col.names = c("Population gate", "Reason", "N excluded"),
    caption   = "Cohort derivation exclusion registry — T2DM-GLP1-RWE-001"
  )
```

```{r waterfall-plot}
ops <- lg_operations() |> filter(op_type == "FILTER")

waterfall <- data.frame(
  label = c("Claims pool", ops$description),
  n     = c(n_pool, ops$rows_out),
  stringsAsFactors = FALSE
)
waterfall$label_short <- c(
  "Claims pool", "Age 40-89", "T2DM diagnosis", "12m lookback",
  "New user", "No insulin", "No ESRD", "12m follow-up"
)
waterfall$step_num <- seq_len(nrow(waterfall))

ggplot(waterfall, aes(x = reorder(label_short, -step_num), y = n)) +
  geom_col(fill = "#1a56db", alpha = 0.8, width = 0.65) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 3.2, colour = "#0f1117") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title    = "Cohort Derivation Attrition",
    subtitle = "T2DM-GLP1-RWE-001 | New-user active comparator design",
    x = NULL, y = "Patients remaining"
  ) +
  theme_minimal(base_size = 11)
```

---

## CONSORT-style disposition

`lg_disposition()` summarises the exclusion registry automatically, derived
from every `lg_filter()` call above — there is no separate group definition
to maintain.

```{r disposition}
lg_disposition(by = "population")
```

---

## Trace a patient

```{r trace}
example_pat <- cohort_final$PATID[1L]
lg_trace(example_pat)
```

```{r trace-excluded}
excluded_pat <- claims_raw$PATID[claims_raw$T2DM_DX == FALSE][1L]
lg_trace(excluded_pat)
```

---

## Outcome analysis

```{r outcome}
outcome_summary <- cohort_final |>
  group_by(INDEX_DRUG) |>
  summarise(
    N          = n(),
    Events     = sum(MACE_FLAG),
    PersonYrs  = round(sum(pmin(FOLLOW_MO, 12) / 12), 1),
    Rate_100py = round(Events / PersonYrs * 100, 2),
    .groups    = "drop"
  )

knitr::kable(
  outcome_summary,
  col.names = c("Treatment", "N", "MACE events", "Person-years", "Rate per 100 PY"),
  caption   = "Primary outcome: MACE at 12 months — unadjusted"
)
```

```{r outcome-plot}
outcome_summary |>
  ggplot(aes(x = INDEX_DRUG, y = Rate_100py, fill = INDEX_DRUG)) +
  geom_col(width = 0.5, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.2f\nper 100 PY", Rate_100py)),
            vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("GLP1RA" = "#1a56db", "DPP4i" = "#6b6f80")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title    = "Crude MACE Rate by Treatment Arm",
    subtitle = "T2DM-GLP1-RWE-001 | Unadjusted, pre-PS matching",
    x = NULL, y = "MACE events per 100 person-years"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

## Provenance report

```{r report, eval=FALSE}
lg_report(
  output   = "outputs/T2DM-GLP1-RWE-001_cohort_provenance_v1.html",
  title    = "Cohort Derivation Provenance — T2DM-GLP1-RWE-001",
  study_id = "T2DM-GLP1-RWE-001",
  sponsor  = "Example Pharma Ltd",
  author   = "J. Smith, Biostatistician"
)
```

---

## End session

```{r end}
lg_end()
```

---

## Why row-level provenance matters in RWE

Aggregate exclusion flow charts are table stakes. Row-level provenance is
what distinguishes a reproducible RWE study from a defensible one:

**For replication:** another analyst can independently derive the same cohort and verify that the same patients are included/excluded at each step — not just the same counts.

**For regulatory review:** NICE, ICER, and FDA RWE reviewers increasingly ask for evidence that the operational cohort definition matches the pre-specified SAP. Row-level traceability makes this verifiable, not just assertable.

**For sensitivity analyses:** re-running the derivation under an alternative lookback window or relaxed eligibility criterion, `lg_trace()` shows exactly which patients changed status and why.

**For audit:** the provenance report is a self-contained HTML document recording the complete derivation history — without requiring access to the original code or data environment.

---

## Session information

```{r session-info}
sessionInfo()
```
